### 2024-11-29 teaching/math240f24/draft.tex
lecture 28 nov

\def\datestamp#1#2 {\llap{{\oldstyle #1}{\rm.}{\sc #2}\hskip12pt}}
\def\bar{\overline}
\def\symdiff{\mathrel{\triangle}}
\def\To{\Rightarrow}
\def\Gets{\Leftarrow}
\font\titlefont=cmbx12 at 14 pt
\def\cont#1#2{#1\hskip-2pt\leaders\hbox to 8pt{\hss.\hss}\hfil\hbox to 1em{\hss#2}}
\def\divides{\mid}
\def\percent{\mathrel{\%}}
\def\norm#1{|\!|#1|\!|}
\def\dist{\op{\rm dist}}

@@

Let $(a_n) = a_0, a_1, a_2, \ldots$ be a sequence of real numbers (in all
the examples we consider, they will be integers). A {\it recurrence relation}
for the sequence is a formula for $a_n$ in terms of some of the $a_m$ with $m<n$.
For instance, the {\it Fibonacci sequence} is the sequence $F_n$ defined by $F_0 = 0$,
$F_1 = 1$, and $F_n = F_{n-1} + F_{n-2}$ for $n\ge 1$. The first few terms of $F_n$ are
$$ 0 , 1, 1, 2, 3, 5, 8, 13, 21, \ldots.$$
The sequence is named for the Italian mathematician Fibonacci, who in his 1202 book {\sl Liber Abaci}
uses these numbers to describe the growth of a population of rabbits under certain idealised
conditions. The sequence was known to Indian linguists many centuries earlier, as it pertains
to the number of different ways to arrange long and short Sanskrit syllables into lines of poetry.
\bye

\medskip\boldlabel Finding recurrence relations.
First we show an example of how to find a recurrence relation for a given sequence.
For $n\ge 1$, let $C_n$ be the number of binary strings of length $n$ not containing $00$.
We have $C_1 = 2$ ($0$ and $1$ are both fine) and $C_2 = 3$ ($01$, $10$, and $11$ are all fine,
but $00$ is not).
Now let $n\ge 2$ and consider how to derive a count for $C_n$, assuming
we know the smaller values of $n$. Consider a string of length $n$ with no $00$.
Either the string ends in $1$ or it ends in $10$, since if it ends with a $0$ the second-last
bit must be $1$. In the first case, there are $n-1$ bits preceding the $1$, and the only requirement
is that this string of $n-1$ bits not contain $00$. So there are $C_{n-1}$ possibilities in the
first case. In the second case, there are $n-2$ bits preceding the $10$, and the requirement
again is that the string $n-2$ bits not contain $00$. Hence in the second case there are
$C_{n-2}$ possibilities. Adding up the two possibilities, we have
$$C_n = C_{n-1} + C_{n-2}.$$
This is the exact same recurrence relation as for the Fibonacci numbers! The only difference
is that the answer is shifted (e.g., $C_1 = 3 = F_3$). So in general, $C_n = F_{n+2}$
for all integers $n\ge 1$.

@@

\medskip\boldlabel\datestamp{26}{} Non-homogeneous recurrences of degree 1.
So far, the only recurrences we know how to solve are
linear homogeneous recurrences of degree $1$.
As a step up in difficulty, let us now consider non-homogeneous recurrences of degree $1$.

@@

\medskip\boldlabel Recurrences with constant coefficients.
We now turn to the case where the $f_i$ are all constant functions. This means
there are constants $c_1, c_2, \ldots, c_k$ such that $f_1(n) = c_1$ for all $n$,
$f_2(n) = c_2$ for all $n$, and so on.
In light of Theorem~{\thmnonhomo}, we can
restrict our attention to homogeneous recurrences, since we can solve non-homogeneous
recurrences by solving their associated homogeneous recurrences and then applying the theorem.
Homogeneous recurrences with constant coefficients look like
$$a_n = c_1 a_{n-1} + c_2 a_{n-2} + \cdots + c_k a_{n-k},$$
for some degree $k\ge 1$. We define the {\it characteristic equation} of such a recurrence to be the
equation
$$x^k = c_1 x^{k-1} + c_2 x^{k-2} + \cdots + c_k,$$
and the {\it characteristic polynomial} is the polynomial obtained by subtracting
the right-hand side from the left-hand side:
$$x^k - c_1 x^{k-1} + c_2 x^{k-2} + \cdots + c_k$$
It turns out that this
polynomial entirely determines the general solution to the recurrence. We shall state
and prove this theorem in the special case that $k=2$ and the two roots of the characteristic
polynomial are distinct.

@@

>>>>

\def\datestamp#1#2 {\llap{{\oldstyle #1}{\rm.}{\sc #2}\hskip12pt}}
\def\bar{\overline}
\def\symdiff{\mathrel{\triangle}}
\def\To{\Rightarrow}
\def\Gets{\Leftarrow}
\font\titlefont=cmbx12 at 14 pt
\def\cont#1#2{#1\hskip-2pt\leaders\hbox to 8pt{\hss.\hss}\hfil\hbox to 2em{\hss#2}}
\def\divides{\mid}
\def\percent{\mathrel{\%}}
\def\norm#1{|\!|#1|\!|}
\def\dist{\op{\rm dist}}

@@

Let $(a_n) = a_0, a_1, a_2, \ldots$ be a sequence of real numbers (in all
the examples we consider, they will be integers). A {\it recurrence relation}
for the sequence is a formula for $a_n$ in terms of some of the $a_m$ with $m<n$.
For instance, the {\it Fibonacci sequence} is the sequence $F_n$ defined by $F_0 = 0$,
$F_1 = 1$, and $F_n = F_{n-1} + F_{n-2}$ for $n\ge 2$. The first few terms of $F_n$ are
$$ 0 , 1, 1, 2, 3, 5, 8, 13, 21, \ldots.$$
The sequence is named for the Italian mathematician Fibonacci, who in his 1202 book {\sl Liber Abaci}
used these numbers to describe the growth of a population of rabbits under certain idealised
conditions. The sequence was known to Indian linguists many centuries earlier, as it pertains
to the number of different ways to arrange long and short Sanskrit syllables into lines of poetry.

\medskip\boldlabel\datestamp{28}{xi} Finding recurrence relations.
First we show an example of how to find a recurrence relation for a given sequence.
For $n\ge 1$, let $C_n$ be the number of binary strings of length $n$ not containing $00$.
We have $C_1 = 2$ ($0$ and $1$ are both fine) and $C_2 = 3$ ($01$, $10$, and $11$ are all fine,
but $00$ is not).
Now let $n\ge 2$ and consider how to derive a count for $C_n$, assuming
we know the smaller values of $n$. Consider a string of length $n$ with no $00$.
Either the string ends in $1$ or it ends in $10$, since if it ends with a $0$ the second-last
bit must be $1$. In the first case, there are $n-1$ bits preceding the $1$, and the only requirement
is that this string of $n-1$ bits not contain $00$. So there are $C_{n-1}$ possibilities in the
first case. In the second case, there are $n-2$ bits preceding the $10$, and the requirement
again is that the string $n-2$ bits not contain $00$. Hence in the second case there are
$C_{n-2}$ possibilities. Adding up the two possibilities, we have
$$C_n = C_{n-1} + C_{n-2}.$$
This is the exact same recurrence relation as for the Fibonacci numbers! The only difference
is that the answer is shifted (e.g., $C_1 = 3 = F_3$). So in general, $C_n = F_{n+2}$
for all integers $n\ge 1$.

@@

\medskip\boldlabel Non-homogeneous recurrences of degree 1.
So far, the only recurrences we know how to solve are
linear homogeneous recurrences of degree $1$.
As a step up in difficulty, let us now consider non-homogeneous recurrences of degree $1$.

@@

\medskip\boldlabel\datestamp{26}{} Recurrences with constant coefficients.
We now turn to the case where the $f_i$ are all constant functions. This means
there are constants $c_1, c_2, \ldots, c_k$ such that $f_1(n) = c_1$ for all $n$,
$f_2(n) = c_2$ for all $n$, and so on.
In light of Theorem~{\thmnonhomo}, we can
restrict our attention to homogeneous recurrences, since we can solve non-homogeneous
recurrences by solving their associated homogeneous recurrences and then applying the theorem.
Homogeneous recurrences with constant coefficients look like
$$a_n = c_1 a_{n-1} + c_2 a_{n-2} + \cdots + c_k a_{n-k},$$
for some degree $k\ge 1$. We define the {\it characteristic equation} of such a recurrence to be the
equation
$$x^k = c_1 x^{k-1} + c_2 x^{k-2} + \cdots + c_k,$$
and the {\it characteristic polynomial} is the polynomial obtained by subtracting
the right-hand side from the left-hand side:
$$x^k - c_1 x^{k-1} + c_2 x^{k-2} + \cdots + c_k$$
It turns out that this
polynomial entirely determines the general solution to the recurrence. We shall state
and prove this theorem in the special case that $k=2$ and the two roots of the characteristic
polynomial are distinct.

@@

%\midinsert
%$$\epsfbox{bijection.ps}$$
%\vskip5pt
%\caption{A planted plane tree with root label $5$ and its corresponding walk in $T_4$.}
%\endinsert
%\goodbreak
